---
title: "NBA Team Strength"
author: "Nick Kim, Jia Guo, Suvro Mukherjee"
date: "2025-04-18"
output:
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: true
---

![Rockets SLAM Cover (Slam 254, 2025)](https://drive.google.com/uc?export=view&id=1pIHphgVGkqYD_Y30jos5B0bPQg4v1u3O)

## Introduction

Basketball, like all sports, is wildly unpredictable. A game filled with runs of hot and cold streaks, any team can win on a given night. However, in all this chaos is there is there an underlying truth? Are some teams just flat out better than another, and can we measure by how much? Luckily there is a traditional method to try and capture such a state, ELO. This simple and beautiful statistic, introduced by Arpad Elo, sets all teams or players at the same base rating and over time adjusts their score depending on the result of their game and how strong their opponent is (Wikipedia, 2025). However, this simple approach fails to account for the randomness of each event. A team could be having on off night, a crucial player could be injured or traded, all of which would alter their ELO rating on that given day. In this report we'll introduce what we'll call POMP-ELO which will try and remedy these concerns by focusing on modeling the team strength for the Houston Rockets. We'll walk you through our data preparation, our model selection as well as comparison to other baseline models, and finally a conclusion on our results. 

## Data Preparation 

While ELO is a wildly known metric there is no readily available data for tracking a teams ELO and their respective opponents ELO throughout a season. For this reason we had to calculate this ourselves by first pulling all matchups in the 2023-24, and 2024-25 season from the database Basketball Reference. Using the scores and results of each game we then use the formula of ELO to calculate this metric for every team:

$$
E_{S} = \frac{1}{1+10^{\frac{TS - OP}{400}}}
$$
$$
TS = TS \pm K\cdot E_{S}
$$
Where $TS$ is team strength and $OP$ and opponent strength. Then will update $TS$ depending on the result of the game, adding if win and subtracting if loss. Finally, the optimal K-factor, how sensitive ELO adjusts to new results, for basketball is found to be 20 which is what we'll use as well (Australia Sports Tipping, 2025).

```{r include=FALSE}
library(readxl)
matchups <- read_excel("/Users/nicholaskim/Documents/STAT-531/final/data/matchups.xlsx")
```

Code for ELO 
```{r}
update_elo <- function(rating1, rating2, result, K = 20) {
  E1 <- 1 / (1 + 10^((rating2 - rating1) / 400))  # Expected score for team 1
  E2 <- 1 / (1 + 10^((rating1 - rating2) / 400))  # Expected score for team 2
  
  if (result == 1) {  # Team 1 wins
    rating1_new <- rating1 + K * (1 - E1)
    rating2_new <- rating2 + K * (0 - E2)
  } else {  # Team 2 wins
    rating1_new <- rating1 + K * (0 - E1)
    rating2_new <- rating2 + K * (1 - E2)
    
  } 
  
  return(c(rating1_new, rating2_new))
}
```

```{r include=FALSE}
elo_ratings <- data.frame(Team = unique(c(matchups$Visitor, matchups$Home)),
                          Elo = rep(1500, length(unique(c(matchups$Visitor, matchups$Home)))))

matchup23 <- matchups[matchups$Season>=2023,]
matchup23$Result <- ifelse(matchup23$PTS_V > matchup23$PTS_H, 1, 2)  

elo_v <- c()
elo_h <- c()

# Loop through matchups to update Elo ratings
for (i in 1:nrow(matchup23)) {
  team1 <- matchup23$Visitor[i]
  team2 <- matchup23$Home[i]
  
  # Get current ratings using match()
  rating1 <- elo_ratings$Elo[match(team1, elo_ratings$Team)]
  rating2 <- elo_ratings$Elo[match(team2, elo_ratings$Team)]
  
  # Update Elo ratings
  updated_ratings <- update_elo(rating1, rating2, matchup23$Result[i])
  
  # Save updated ratings back into the data frame
  elo_ratings$Elo[elo_ratings$Team == team1] <- updated_ratings[1]
  elo_ratings$Elo[elo_ratings$Team == team2] <- updated_ratings[2]
  
  elo_v <- c(elo_v,updated_ratings[1])
  elo_h <- c(elo_h,updated_ratings[2])
  
}

matchup23$ELO_V <- elo_v
matchup23$ELO_H <- elo_h

r_23 <- matchup23[(matchup23$Visitor=="Houston Rockets")|(matchup23$Home=="Houston Rockets"),]

n <- nrow(r_23)
rockets_elo_df <- data.frame(date = as.Date(rep(NA, n)), elo = rep(NA, n), opp_elo = rep(NA, n))
row_counter <- 1

for (i in 1:n) {
  if (r_23[i, 3] == "Houston Rockets") {
    rockets_elo_df[row_counter, ] <- list(as.Date(r_23[[i, 2]],format = "%a, %b %d, %Y"), as.numeric(r_23[i, 9]), as.numeric(r_23[i,10]))
    row_counter <- row_counter + 1
  } else if (r_23[i, 5] == "Houston Rockets") {
    rockets_elo_df[row_counter, ] <- list(as.Date(r_23[[i, 2]],format = "%a, %b %d, %Y"), as.numeric(r_23[i, 10]),
                                          as.numeric(r_23[i,9]))
    row_counter <- row_counter + 1
  }
}

# Remove unused rows
rockets_elo_df <- rockets_elo_df[1:(row_counter - 1), ]
```

We now have the data for our latent space which we'll try to model with our POMP model. To do this we'll use the following covariates: Average Last 5 game Total BPM, Home, Opponent ELO, and Average Last 5 Opponent Total BPM (depending on the POMP model we chose). These stats were similarily pulled from Basketball Reference using the Box Score statistics for each game.

Here BPM is a statistics that trys to measure how impactful a certain player was in a given game. For the purposes of our model we added the BPM's for all the starters (5 players that started the game) to get a measure of Total BPM. The rational behind this is that the starters are generally the best players for each team thus this measure would give us the best representation of how good a team performed on a given night. To take one step further we then took the average of these Total BPM's in their last 5 games to get a measure of team momentum. In the case, if the number of games are less than 5 we then took the average with repsect to how many games they played up until that point. This is more clearly seen when we take a look at the data: 

```{r include=FALSE}
bpm <- read_excel("/Users/nicholaskim/Documents/STAT-531/final/data/BPM.xls")
bpm["elo"] <- rockets_elo_df$elo
bpm["time"] <- rockets_elo_df$time
bpm["opp_elo"] <- rockets_elo_df$opp_elo
```

```{r echo=FALSE}
bpm[1:10,c('Date','Total BPM','Last 5 Games BPM')]
```

We'll introduce two different POMP models were two different assumptions were made: 
1. Opp ELO is a covariate 
2. Opp ELO is a state itself
This will become more clear in our POMP section but under the 2nd condition we'll use the Average Last 5 BPM for the opponent to then adjust the Opponent strength while under the 1st it'll not fluctuate is a random process like team strength. 

## Baseline Models

## Linear Regression 

## ARIMA

## POMP 

## Model Comparison

## Conclusion 

## References


